Demographic inference using the SFS with \moments and \demes

### 2025-03-26 chapter/paper.tex
it's not good writing, but it's writing

\begin{itemize}
    \item to run inference, need to specify parameters to be fit
    \item the parameters/options file points to values in the demographic model
        to be fit, along with bounds on those parameters and any constraints
        among parameters
    \item this options file is also a YAML -- see snippets of examples below, and
        full details are given in the above URL
\end{itemize}

@@

Using the split-with-migration model defined above, we simulated data for 30
diploid individuals from both populations using \msprime
\citep{baumdicker2022efficient}. These simulations consided of 500 regions,
each of length 1 Mb, with per-base recombination and mutation rates of
$10^{-8}$. The joint SFS was computed using \tskit's
\texttt{ts.allele\_frequency\_spectrum()} function separately on each replicate
region and summing across regions to find the total SFS across 500 Mb of data
(Figure~\ref{fig:im}B--D).

>>>>

In order to optimize parameters in a \demes model, we need a way to specify
which parameters should be fit. For this, we use a separate YAML-formatted file
to define \texttt{parameters}, each of which points to the value(s) in the
demographic model and sets lower and upper bounds. This file also specifies any
relative \texttt{constraints} between pairs of parameters, to ensure the
optimization routine only explores valid model space. Below, we have included
a truncated parameters file -- the full example is available on GitHub.
\begin{minted}[gobble=4, frame=single]{YAML}
    parameters:
    - name: T0
      values:
      - demes:
          ancestral:
            epochs:
              0: end_time
      lower_bound: 0
      upper_bound: 2e5
    - name: T1
      values:
      - demes:
          ancestral:
            epochs:
              1: end_time
      lower_bound: 0
      upper_bound: 1e6
    - name: Ne
      values:
      - demes:
          ancestral:
            epochs:
              0: start_size
      lower_bound: 1e2
      upper_bound: 1e6
    ...
    constraints:
    - params: [T0, T1]
      constraint: greater_than
\end{minted}

Taking the demographic model, parameters file, and observed SFS together, we
can perform inference using the \texttt{moments.Demes.Inference.optimize(...)}
function.  The initial parameter guesses are given by the input demographic
model, which may be perturbed using a keyword argument. Any value in the
demographic model that is not specified in the parameters file will remain
unchanged, and is therefore treated as a fixed parameter. Here, we also assume
we have an estimate for the total mutation rate, $U$.
\begin{minted}[gobble=4, frame=single]{YAML}
    import moments

    data = moments.Spectrum.from_file('data.fs')
    U = 1e-8 * 5e8  # per-base mutation rate times the total length

    model_file = 'model.yaml'
    params_file = 'params.yaml'
    output = 'model.fit.yaml'

    ret = moments.Demes.Inference.optimize(
        model_file, params_file, data,
        perturb=1, uL=U, output=output
    )
\end{minted}

Confidence intervals may also be computed, once a (locally) optimal model has
been found.

@@

Using the split-with-migration model defined above (illustrated in
Figure~\ref{fig:im}A), we simulated data for 30 diploid individuals from both
populations using \msprime \citep{baumdicker2022efficient}. These simulations
consided of 500 regions, each of length 1 Mb, with per-base recombination and
mutation rates of $10^{-8}$. The joint SFS was computed using \tskit's
\texttt{ts.allele\_frequency\_spectrum()} function separately on each replicate
region and summing across regions to find the total SFS across 500 Mb of data
(Figure~\ref{fig:im}B--D).

We fit three demographic models to the simulated data. In addition to
reinferring the parameters from the simulated model, we fit two simpler models
to the same data. These had fewer parameters, both omitting the size change
deeper in time in the ancestral population. This is meant to crudely mimic the
scenario that we often face, in which the true history is more complicated than
the parameterized model. It also highlights biases in inferences that can arise
when features of the true history are not included.

When fitting the two misspecified models that do not allow for size changes in
the ancestral population, this split time is inferred to be substantially
deeper in the past than the true split time. This effect, as demonstrated by
\citet{momigliano2020biases}, is due to the decreased coalescence rate within
the ancestral population between the time of the expansion and divergence of
the descendent populations. These two misspecified models
(Figure~\ref{fig:im}H,I) also provide a worse fit to the data, as seen by the
lower log-likelihoods and the large residuals between model predictions and
data (Figure~\ref{fig:im}E-G).
